Safe Memory Regions for Big Data Processing

### 2016-06-03 paper/type-inference.tex
typeinference

\paragraph{Terminology.}
In the sequel, we will be using \emph{region identifiers} in four different roles.
A region identifier may serve as
(a) A \emph{formal region parameter} of a class or method, or
(b) a \emph{static region identifier} introduced by a \C{letregion} construct, or
(c) an  \emph{open transferable region identifier} introduced by an \C{open} construct, or
(d) a \emph{region variable}, introduced to represent an unknown actual region parameter
of a method invocation or object alloation,
which will be bound to a region identifier of one of the three
preceding kinds by the end of the type inference.

Fig.~\ref{fig:type-inference-algo} presents the high-level outline of the type
inference algorithm.
The algorithm consists of the following steps:
\begin{enumerate}
 \item The first step elaborates the input program by introducing \emph{(formal and actual) region parameters},
and \emph{region variables} (representing yet undetermined \emph{actual region parameters})
as follows:
\begin{enumerate}
\item For every class determine a set of \emph{region parameters} ($\rho$)
       it should be parametrized over.
\item For every field of the class, determine its region parametrized type by
       generating as many fresh \emph{actual region parameters} as required by its class (type),
       which in turn become formal region parameters of the containing class.
\item  For every method and function type, determine its corresponding region
   type template by introducing fresh region parameters as required by
   the type of each parameter and return value (in addition to the allocation region
   parameter).
 \item Elaborate every \C{new} expression, method call or function application by
   introducing \emph{fresh region variables} (as required by the region type
   template of the called method or function).
\item Introduce for each class and method, a \emph{predicate variable} ($\varphi$) to denote
   an undetermined set of constraints over the region parameters of that
   class/method.
\end{enumerate}
 \item In the second step, we analyze the program to generate a set of constraints that capture
   well-formedness requirements and subtyping relationships between type templates
   that must hold (as per the static semantics in Fig.~\ref{fb-staticsem}).
 \item We solve the generated set of constraints using our fixpoint constraint
   solving algorithm \csolvestar, which reduces the constraint solving problem to
   an abduction problem. 
% in a Herbrand constraint system, and then relies on \csolve,
% our abduction solver for that domain. 
 \item If the solver succeeds, it returns substitution functions $\substFn_\rho$ and
  $\substFn_\varphi$ for free region and predicate variables, respectively, introduced in the earlier
phases. We apply these substitutions to the elaborated program to produce the final program.
Note that if the original program in $\absof{\FB}$ contains unsafe
references, for example, a reference from a transferable region to a
stack regions, then the constraints generated during the elaboration
are not satisfiable. In such case, $\solve$ fails to solve
constraints, causing the program to be rejected.
\end{enumerate}

>>>>

\paragraph{Overview.}
In the sequel, we will be using \emph{region identifiers} in four different roles.
A region identifier may serve as
(a) A \emph{formal region parameter} of a class or method, or
(b) a \emph{static region identifier} introduced by a \C{letregion} construct, or
(c) an  \emph{open transferable region identifier} introduced by an \C{open} construct, or
(d) a \emph{region variable}, introduced to represent an unknown actual region parameter
of a method invocation or object alloation,
which will be bound to a region identifier of one of the three
preceding kinds by the end of the type inference.

Fig.~\ref{fig:type-inference-algo} presents the high-level outline of the type
inference algorithm.
The algorithm consists of the following steps:
\begin{enumerate}
 \item The first step elaborates the input program by introducing \emph{formal region parameters}
   (for each class and method), and \emph{region variables} (representing yet undetermined
   \emph{actual region parameters}). We also introduce for each class and method, a
   \emph{predicate variable} ($\varphi$) to denote an undetermined set of constraints
   over the region parameters of that class/method.

%%    as follows:
%% \begin{enumerate}
%% \item For every class determine a set of \emph{region parameters} ($\rho$)
%%        it should be parametrized over.
%% \item For every field of the class, determine its region parametrized type by
%%        generating as many fresh \emph{actual region parameters} as required by its class (type),
%%        which in turn become formal region parameters of the containing class.
%% \item  For every method and function type, determine its corresponding region
%%    type template by introducing fresh region parameters as required by
%%    the type of each parameter and return value (in addition to the allocation region
%%    parameter).
%%  \item Elaborate every \C{new} expression, method call or function application by
%%    introducing \emph{fresh region variables} (as required by the region type
%%    template of the called method or function).
%% \item Introduce for each class and method, a \emph{predicate variable} ($\varphi$) to denote
%%    an undetermined set of constraints over the region parameters of that
%%    class/method.
%% \end{enumerate}

 \item In the second step, we analyze the program to generate a set of constraints
   (over the region identifiers and the predicate variables)
   that must hold (as per the static semantics in Fig.~\ref{fb-staticsem}).

 \item We solve the generated set of constraints using our fixpoint constraint
   solving algorithm \csolvestar, which reduces the constraint solving problem to
   an abduction problem. If the original program in $\absof{\FB}$ contains unsafe
   references, for example, a reference from a transferable region to a
   stack region, then the constraints generated during the elaboration
   are not satisfiable. In such a case, \csolvestar{} fails to solve
   the constraints.
% in a Herbrand constraint system, and then relies on \csolve,
% our abduction solver for that domain.

 \item If the solver succeeds, it returns substitution functions $\substFn_\rho$ and
  $\substFn_\varphi$ for free region and predicate variables, respectively, introduced in
  step 1. We apply these substitutions to the elaborated program to produce the final program.

\end{enumerate}
